---
title: "Capelin and Cod Distribution and Abundance in Barents Sea in 2013 (using Fall et al. 2018)"
author: "Janelle Morano"
date: "last updated: 4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Northeast Arctic cod (_Gadus morhua_) and Barents Sea capelin (_Mallotus villosus_) are important commercial fisheries stocks in the Barents Sea off the coast of Norway. Cod is a generalist predator, but has a preference for capelin and is the main predator on post-larval capelin. 

The distribution of cod and capelin varies seasonally, and the seasonal predation of cod on capelin follows. The northern Barents Sea is seasonally ice-covered. When the ice melts and recedes in the spring, zooplankton production blooms, and the Capelin migrate northward from the Norwegian coast to feed. Cod follow after (feeding on capelin and other things?). In early fall, as the zooplankton production decreases, both species begin migrating back south. Cod predation on capelin is higher in the winter than the summer (is this due to availability of other prey?). 

Both species spawn in early spring. Capelin spawns along the northern coast of Norway and Russia, and
cod’s main spawning ground is further south along the Norwegian coast in the Lofoten area. 

Cod abundance has been increasing, and their distribution has been expanding northward, potentially impacting capelin. The Norwegian stock assessment model relies on several assumptions related to the seasonal interaction between cod and capelin, but does not explicitly incorporate spatial distribution and overlap of the two species. Additionally, in the last couple of decades, water tempteratures increased, cod biomass doubled, and the capelin stock collapsed twice.

The purposed of this study is to explore the spatial distribution of cod and capelin in one year, given available data from Fall et al. 2018. The authors modeled the spatial distribution and abundance of cod and capelin, and the seasonality of these patterns from 2004-2015. Here, I am using the available data to explore the patterns of spatial distribution, abundance, and overlap of cod and capelin in the autumn of 2013. I will apply multiple spatial analysis methods to understand the approaches, interpretations, and limitations, and look for how to expand the analyses to address temporal changes.

## Research Question
Research Question: How do capelin and cod overlap in the Barents Sea?

### Objectives:

1. What is the abundance and distribution of capelin and cod across the Barents Sea?
1. How much do they overlap? Are there patterns in overlap?
1. Are there correlations with water temperature or depth?
1. Reconstruct, in part, the methods and results of Fall et al. 2018 to be able to understand and explore further, the Barents Sea ecosystem and general approaches to the spatial distribution of fishereies.

## Methods

### Data
Data are from the Norwegian ecosystem survey (2004-2015) covering the whole shelf in the main feeding season. These are some of the data by Fall et al. 2018 and available from [Dryad Digital Repository](https://doi.org/10.5061/dryad.pv3rc1m). The ecosystem survey occurs (August-October) and is a stratified-random bottom trawl survey, sampling at fixed locations in the Barents Sea and around Svalbard. A Campelen demersal shrimp trawls are equipped with Simrad EK60 echo sounders to integrate acoustic backscatter sampling into swept area abundance estimates. CTD and depth measurements are also done on the trawl.

Cod densities (number of individuals/nmi^2) were estimated using standard methods for
cod swept area calculation in the Barents Sea, that is, number caught at each trawl station divided by trawled area, assuming that the effective fishing width along the trawled transect
is dependent on cod length

Data are acoustic backscatter of capelin and total density of immature and mature cod (ecosystem survey: mature cod >75 cm), bottom depth (m), bottom temperature (Celsius), and pelagic temperature (Celsius).

### Biological and Environmental Factors

* Correlate among capelin, immature cod, and mature cod densities and with bottom depth and water temperature with linear regression.
* Repeat with GAM (maybe)

### Spatial Analyses

* Kriging capelin, immature cod, and mature cod densities across survey area.
* INLA (maybe)

```{r methods, echo=FALSE, results='hide'}

#Lines 7-39,66-67 code from Capelin-Cod-BarentsSea-2013.R are copied here. Projection is eliminated here.

##########
# Setup
##########
# Working directory
setwd("/Users/janellemorano/Git/Reference-R-scripts/Spatial-Stats/Cod-Capelin-Project")

# Libraries
library(sp)
library(rgdal)
library(gstat)
library(geoR)
library(RColorBrewer)
library(classInt)

#All of the data from Fall et al. 2018
data = read.table("Norwegian part of ecosystem survey cod-capelin 2004-2015.txt", sep = " ",header = T)
head(data)
dim(data)
#2455   42

#make each year a unique number for reference
iyear = unique(data$year)
#assign the lat and lon
data$Xloc = jitter(data$lon)
data$Yloc = jitter(data$lat)
coordinates(data)=c("Xloc","Yloc")

#Make a dataset for just 2013 to simply things.
#######
yy = 2013
datai = data[data$year==yy,]
dim(datai)
#217  42

## Plot survey locations
plot(lat~lon,data=datai)

```

The data were log-transformed to account for the distribution of capelin abundance.

The variogram was fit using a spherical model in gstat in R.

```{r variogram, echo=FALSE, results='hide'}

#Lines 193-205 copied here, suppressing thru commenting plots 1,2.

#######################
# Capelin: Empirical Variogram
########################

# Calculate the empirical variogram for Capelin
capelin.vario = variogram(log(capelinA+1)~1,datai,cutoff=20)
#
# # Plot the empirical variogram
# plot(gamma~dist,capelin.vario,
#      ylim=c(0,max(gamma)),type='n',
#      xlab="Distance",ylab="Semivariance",
#      main=paste("Capelin Variogram",yy))
# points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")


#######################
# Capelin: Fit Model By-Eye
########################

my.range = 6.8
my.nugget = 0.1
my.psill = 6-my.nugget
#
capelin.eye = vgm(model="Sph",psill=my.psill,range=my.range,nugget=my.nugget)
# plot(gamma~dist,capelin.vario,
#      ylim=c(0,max(gamma)),type='n',
#      xlab="Distance",ylab="Semivariance",
#      main=paste("Capelin Variogram By-Eye",yy))
# points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")
# vgmline = variogramLine(capelin.eye,max(capelin.vario$dist))
# lines(gamma~dist,vgmline,lwd=2)

#######################
# Capelin: Fit Actual Model
########################
capelin.fit=fit.variogram(capelin.vario,
                          vgm(model="Sph",psill=my.psill,range=my.range,nugget=my.nugget),
                          fit.method=1)

# Look at estimates
capelin.fit
capelin.psill=capelin.fit$psill[2]
capelin.range=capelin.fit$range[2]
capelin.nugget=capelin.fit$psill[1]

# Plot the data, model and parameter estimates
#####
plot(gamma~dist,capelin.vario,
     ylim=c(0,max(gamma)),type='n',
     xlab="Distance",ylab="Semivariance",
     main=paste("Capelin Variogram",yy))
points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")
vgmline = variogramLine(capelin.fit,max(capelin.vario$dist))
lines(gamma~dist,vgmline,lwd=2)
#
legend("bottomright",legend = c(
  paste("Psill  =",round(capelin.psill,2)),
  paste("Range  =",round(capelin.range,2)),
  paste("Nugget = ",round(capelin.nugget,2))),
  bty="n")
```

Abundance of capelin was estimated using Krig over a space of length = 40 (what does this convert to??)

```{r krig, echo=FALSE}
#Lines 254-286

#######################
# Capelin: Predict: Kriging
########################

# Given the fitted model, let's make some predictions about the data.

# Create a grid of points to predict over
capelin.grid = expand.grid(
  Xloc=seq(min(datai$Xloc),max(datai$Xloc),length=40),
  Yloc=seq(min(datai$Yloc),max(datai$Yloc),length=40))
names(capelin.grid)=c("Xloc","Yloc")
coordinates(capelin.grid)=c("Xloc","Yloc")
capelin.grid = as(capelin.grid, "SpatialPixels")

# Now plot the data and overlay the prediction grid
plot(Yloc~Xloc,capelin.grid,cex=1.2,pch='+',col="green")
points(Yloc~Xloc,datai,pch=".")	

# Predict the value at all the points in the domain
date()	
capelin.ok = krige(log(capelinA+1)~1, datai, capelin.grid, m=capelin.fit)	
date()

```

Need to fix the legend.

## Results

Abundance of cod and capelin over Barents Sea in 2013.

```{r abundance, echo=FALSE, results="hide"}
#Lines 84-160

# Plot capelin and cod relative densities, separately
#####
## Plot relative capelin densities for 2013
# Set colors
pal = brewer.pal(5,"Blues")
q5 = classIntervals(datai$capelinA, n=5, style="quantile")
q5Colors = findColours(q5,pal)

# Plot
plot(c(min(datai$Xloc),max(datai$Xloc)),
     c(min(datai$Yloc),max(datai$Yloc)),
     xlab="Longitude",ylab="Latitude",type="n")
points(datai,col=q5Colors,pch=1,add=T, cex = 2*(datai$capelinA/max(datai$capelinA)))
legend("bottomleft",fill=attr(q5Colors,"palette"),	legend = names(attr(q5Colors,"table")),bty="n")
title(paste("Capelin Abundance",yy))

## Plot relative immature cod densities for 2013
# Set colors
pal2 = brewer.pal(5,"Oranges")
q52 = classIntervals(datai$cod.imm, n=5, style="quantile")
q5Colors2 = findColours(q52,pal2)
plot(c(min(datai$Xloc),max(datai$Xloc)),
     c(min(datai$Yloc),max(datai$Yloc)),
     xlab="Longitude",ylab="Latitude",type="n")
#plot(datai,col=q5Colors2,pch=19,add=T)
points(datai,col=q5Colors2,pch=1,add=T, cex = 2*(datai$cod.imm/max(datai$cod.imm)))
legend("bottomleft",fill=attr(q5Colors2,"palette"), legend = names(attr(q5Colors2,"table")),bty="n")
title(paste("Immature Cod Abundance",yy))

## Plot relative mature cod densities for 2013
# Set colors
pal2.2 = brewer.pal(5,"Greens")
q52.2 = classIntervals(datai$cod.mat, n=5, style="quantile")
q5Colors2.2 = findColours(q52.2,pal2.2)
plot(c(min(datai$Xloc),max(datai$Xloc)),
     c(min(datai$Yloc),max(datai$Yloc)),
     xlab="Longitude",ylab="Latitude",type="n")
#plot(datai,col=q5Colors2,pch=19,add=T)
points(datai,col=q5Colors2.2,pch=1,add=T, cex = 2*(datai$cod.mat/max(datai$cod.mat)))
legend("bottomleft",fill=attr(q5Colors2.2,"palette"), legend = names(attr(q5Colors2.2,"table")),bty="n")
title(paste("Mature Cod Abundance",yy))

#####
## Plot relative capelin, immature and mature cod (TOGETHER) densities for 2013
#####
# And makes the color intervals the same

# Capelin
capelin.pal = brewer.pal(5,"Blues")
capelin.q5 = classIntervals(datai$capelinA, n=5, style="quantile")
capelin.q5Colors = findColours(capelin.q5,capelin.pal)

# Immature Cod
cod.imm.pal = brewer.pal(5,"Oranges")
cod.imm.q5 = classIntervals(datai$capelinA, n=5, style="quantile") #relative to capelin
cod.imm.q5Colors = findColours(cod.imm.q5,cod.imm.pal)

# Mature Cod
cod.mat.pal = brewer.pal(5,"Greens")
cod.mat.q5 = classIntervals(datai$capelinA, n=5, style="quantile") #relative to capelin
cod.mat.q5Colors = findColours(cod.mat.q5,cod.mat.pal)

plot(c(min(datai$Xloc),max(datai$Xloc)),
     c(min(datai$Yloc),max(datai$Yloc)),
     xlab="Longitude",ylab="Latitude",type="n")
#add capelin
points(datai,col=capelin.q5Colors,pch=1, cex = 2*(datai$capelinA/max(datai$capelinA))) 
#add immature cod
points(datai,col=cod.imm.q5Colors,pch=1, cex = 2*(datai$cod.imm/max(datai$cod.imm)))
#add mature cod
points(datai,col=cod.mat.q5Colors,pch=1, cex = 2*(datai$cod.mat/max(datai$cod.mat)))

#add legend: this is terrible looking
legend("topright",fill=attr(capelin.q5Colors,"palette"),	legend = names(attr(capelin.q5Colors,"table")),bty="n")
legend("bottomright",fill=attr(cod.imm.q5Colors,"palette"),	legend = names(attr(cod.imm.q5Colors,"table")),bty="n")
legend("bottomleft",fill=attr(cod.mat.q5Colors,"palette"),	legend = names(attr(cod.mat.q5Colors,"table")),bty="n")
title(paste("Capelin (blue), Immature Cod (orange), Mature Cod (green)  Abundance",yy))

```

Abundance of capelin, immature cod, and mature cod across Barents Sea in 2013.

The combined figure is terrible.

```{r predict, echo=FALSE, results="hide"}
#Lines 278-286

# Plot the prediction
plot(lat~lon,data=datai)
image(capelin.ok["var1.pred"],col=rev(heat.colors(4)),add=T)
#contour(capelin.ok["var1.pred"],add=T)
title(paste("Predicted Log(Capelin+1)",yy))
legend("bottomright",legend=c(0,1,2,3),fill=rev(heat.colors(4)),
       bty="n",title="log(Capelin+1)")
points(lat~lon,data=datai)
summary(capelin.ok["var1.pred"])
```

It looks like the kriging area is greater than the survey area. Need to work on.