---
title: "Capelin and Cod Distribution and Abundance in Barents Sea in 2013"
author: "Janelle Morano"
date: "last updated: 4/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Research Question: How do capelin and cod overlap in the Barents Sea?

1. What is the abundance and distribution of capelin and cod across the Barents Sea?
1. How much do they overlap? Are there patterns in overlap


# Methods

Data are abundance (or biomass) of a variety of fish species and plankton from acoustic surveys in Barents Sea by Norwegian Institute of Marine Science, 2004-2013. The units are unknown: could be abundance, 10^9 or biomass in kg (~10^6).

Here, data for cod and capelin in 2013 were used. Most likely, these were collected in September.

```{r methods, echo=FALSE, results='hide'}

#Lines 7-76, code from Capelin-Cod-BarentsSea-2013.R are copied here.

##########
# Setup
##########
# Working directory
setwd("/Users/janellemorano/Git/Reference-R-scripts/Spatial-Stats/Cod-Capelin-Project")

# Libraries
library(sp)
library(rgdal)
library(gstat)
library(geoR)
library(RColorBrewer)
library(classInt)

#All of the data from 2007-2013
data = read.csv("AllData.csv", head = TRUE)
head(data)
dim(data)
#123056     22

#make each year a unique number for reference
iyear = unique(data$year)
#assign the lat and lon
data$Xloc = jitter(data$lon)
data$Yloc = jitter(data$lat)
coordinates(data)=c("Xloc","Yloc")

#Make a dataset for just 2013 to simply things.
#######
yy = 2013
datai = data[data$year==yy,]
dim(datai)
#15571    22

##############################
# Plot: Visualize Survey Data
##############################
# Projection: NEEDS FIXING

# proj4string(data) <- NA_character_ # remove CRS information from lnd
# proj4string(data) <- CRS("+init=epsg:27700") # assign a new CRS
# EPSG <- make_EPSG() # create data frame of available EPSG codes
# EPSG[grepl("WGS 84$", EPSG$note), ]
# lnd84 <- spTransform(data, CRS("+init=epsg:4326")) # reproject
# saveRDS(object = lnd84, file = "data/lnd84.Rds")

# make the UTM cols spatial (X/Easting/lon, Y/Northing/lat)
# use sf package
# library(sf)
# data.SP <- st_as_sf(data, coords = c("Xloc","Yloc"), crs = 4326)
# ## Plot survey locations
# plot(lat~lon,data=data.SP)
# Projection doesn't look different

## Plot survey locations
plot(lat~lon,data=datai)

# Read in boundary layers
library(rgdal)
setwd("/Users/janellemorano/Git/Reference-R-scripts/Spatial-Stats/Cod-Capelin-Project/Barents Sea Shp")

## Read in boundary layers
ogrInfo(dsn=".",layer="iho")
barents = readOGR(dsn=".",layer="iho")
# proj4string the slot which contains the projection information
proj4string(data)
#[1] NA
plot(barents,add=T,lwd=2)

setwd("/Users/janellemorano/Git/Reference-R-scripts/Spatial-Stats/Cod-Capelin-Project")
```

The data were log-transformed to account for the distribution of capelin abundance.

The variogram was fit using a spherical model in gstat in R.

```{r variogram, echo=FALSE, results='hide'}

#Lines 149-192 copied here, suppressing thru commenting plots 1,2.

#######################
# Capelin: Empirical Variogram
########################

# Calculate the empirical variogram for Capelin
capelin.vario = variogram(log(capelin+1)~1,datai,cutoff=20)
#
# Plot the empirical variogram
# plot(gamma~dist,capelin.vario,
#      ylim=c(0,max(gamma)),type='n',
#      xlab="Distance",ylab="Semivariance",
#      main=paste("Capelin Variogram",yy))
# points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")

#######################
# Capelin: Fit Model By-Eye
########################

my.range = 8.8
my.nugget = 0.8
my.psill = 5.2-my.nugget
#
capelin.eye = vgm(model="Sph",psill=my.psill,range=my.range,nugget=my.nugget)
# plot(gamma~dist,capelin.vario,
#      ylim=c(0,max(gamma)),type='n',
#      xlab="Distance",ylab="Semivariance",
#      main=paste("Capelin Variogram By-Eye",yy))
# points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")
# vgmline = variogramLine(capelin.eye,max(capelin.vario$dist))
# lines(gamma~dist,vgmline,lwd=2)

#######################
# Capelin: Fit Actual Model
########################
capelin.fit=fit.variogram(capelin.vario,
                          vgm(model="Sph",psill=my.psill,range=my.range,nugget=my.nugget),
                          fit.method=1)

# Look at estimates
capelin.fit
capelin.psill=capelin.fit$psill[2]
capelin.range=capelin.fit$range[2]
capelin.nugget=capelin.fit$psill[1]

# Plot the data, model and parameter estimates
#####
plot(gamma~dist,capelin.vario,
     ylim=c(0,max(gamma)),type='n',
     xlab="Distance",ylab="Semivariance",
     main=paste("Capelin Variogram",yy))
points(gamma~dist,capelin.vario,cex=2*np/max(np),pch=16,col="lightblue")
vgmline = variogramLine(capelin.fit,max(capelin.vario$dist))
lines(gamma~dist,vgmline,lwd=2)
#
legend("bottomright",legend = c(
  paste("Psill  =",round(capelin.psill,2)),
  paste("Range  =",round(capelin.range,2)),
  paste("Nugget = ",round(capelin.nugget,2))),
  bty="n")
```

Abundance of capelin was estimated using Krig over a space of length = 40 (what does this convert to??)

```{r krig, echo=FALSE}
#Lines 210-232

#######################
# Capelin: Predict: Kriging
########################

# Given the fitted model, let's make some predictions about the data.

# Create a grid of points to predict over
capelin.grid = expand.grid(
  Xloc=seq(min(datai$Xloc),max(datai$Xloc),length=40),
  Yloc=seq(min(datai$Yloc),max(datai$Yloc),length=40))
names(capelin.grid)=c("Xloc","Yloc")
coordinates(capelin.grid)=c("Xloc","Yloc")
capelin.grid = as(capelin.grid, "SpatialPixels")

# Now plot the data and overlay the prediction grid
plot(Yloc~Xloc,capelin.grid,cex=1.2,pch='+',col="green")
points(Yloc~Xloc,datai,pch=".")	

# Predict the value at all the points in the domain
date()	
capelin.ok = krige(log(capelin+1)~1, datai, capelin.grid, m=capelin.fit)	
date()
#With length=40 grid, takes ~19 mins to run
```

# Results

Abundance of cod and capelin over Barents Sea in 2013.

```{r abundance, echo=FALSE, results="hide"}
#####
## Plot relative capelin & cod (TOGETHER) densities for 2013
#####
# And makes the color intervals the same

# Capelin
capelin.pal = brewer.pal(5,"Blues")
capelin.q5 = classIntervals(datai$capelin, n=5, style="quantile")
capelin.q5Colors = findColours(capelin.q5,capelin.pal)

# Cod
cod.pal = brewer.pal(5,"Oranges")
cod.q5 = classIntervals(datai$capelin, n=5, style="quantile") #relative to capelin
cod.q5Colors = findColours(cod.q5,cod.pal)

plot(c(min(datai$Xloc),max(datai$Xloc)),
     c(min(datai$Yloc),max(datai$Yloc)),
     xlab="Longitude",ylab="Latitude",type="n")
points(datai,col=capelin.q5Colors,pch=1, cex = 2*(datai$capelin/max(datai$capelin))) 
# Add cod on top, but then the legend and title need to be fixed
points(datai,col=cod.q5Colors,pch=1, cex = 2*(datai$cod/max(datai$cod)))
legend("topright",fill=attr(capelin.q5Colors,"palette"),	legend = names(attr(capelin.q5Colors,"table")),bty="n")
legend("bottomright",fill=attr(cod.q5Colors,"palette"),	legend = names(attr(cod.q5Colors,"table")),bty="n")
title(paste("Capelin (blue) and Cod (orange) Abundance",yy))

```

Predicted abundance of capelin across Barents Sea in 2013

```{r predict, echo=FALSE, results="hide"}
#Lines 234-242

# Plot the prediction
plot(barents)
image(capelin.ok["var1.pred"],col=rev(heat.colors(4)),add=T)
#contour(capelin.ok["var1.pred"],add=T)
title(paste("Predicted Log(Capelin+1)",yy))
legend("bottomright",legend=c(0,1,2,3),fill=rev(heat.colors(4)),
       bty="n",title="log(Capelin+1)")
plot(barents,add=T)
summary(capelin.ok["var1.pred"])
```
