---
title: "Step-by-Step Spatial Analysis"
author: "Janelle Morano"
date: "2/28/2020"
output: html_document
---
## Geostatiscal Procedure
When you have data that are spatially correlated, often, points that are close together are going to be more similar than points that are farther apart. This is autocorrelation. When you have spatial data, the goal is to find trends in the data that you are interested in understanding while accounting for the  autocorrelation. You don't want it to muddle your interpretations. 

Some terms to remember...

**Variance**: how far a set of random numbers are spread out (deviate) from the average value; the square of the standard deviation.

**Covariance**: the joint variability of two random variables; the tendency in the linear. Random variable whose covariance is zero are uncorrelated, and independent.

**Variance-Covariance Matrix**: square matrix that contains the variances and covariances associated with variables. The diagonal elements contain the variances of the variables; the off-diagonal elements contain the covariances between all possible pairs of the variables.

**Residual**: the difference between each observations and the sample mean.

**Empirical Variogram**: description of the spatial dependence of the data

**Actual Variogram**: model of the spatial dependence of the data.

**Variogram Models**: Matern, Power, Spherical, Exponential, Gaussian, Linear

```{r variogrammodels, echo =FALSE}
library(sp)
library(gstat)
v.psill=100
v.range=10
v.nugget=20
#
# Trellis plot of variogram models
#
v.data=data.frame(rbind(
variogramLine(vgm(v.psill, "Exp", v.range, v.nugget),maxdist=15),
variogramLine(vgm(v.psill, "Sph", v.range, v.nugget),maxdist=15),
variogramLine(vgm(v.psill, "Gau", v.range, v.nugget),maxdist=15),
variogramLine(vgm(v.psill, "Lin", v.range, v.nugget),maxdist=15),
variogramLine(vgm(v.psill/5, "Pow",range=.5,v.nugget),maxdist=15),
variogramLine(vgm(v.psill, "Mat", 1,v.nugget,kappa=.1),maxdist=15)))
#
v.data$model=c(
	rep("Exp",200),rep("Sph",200),rep("Gau",200),
      rep("Lin",200),rep("Pow",200),rep("Mat",200))
#
library(lattice)
xyplot(gamma~dist|model,data=v.data,type="l",lwd=3)
#
```


### Basic Approach

The basic approach to working with geospatial data follows these steps:

1.Plot the locations
    a.  Identify x and y and put it in geographic context
    b.  Plot the measurements at locations
1. Plot the empirical variogram and find the best model and parameters
1. Fit the actual variogram model
1. Make predictions
    
Not sure when these happen:    
* Calculate the variance (matrix of variance)
* Calculate the co-variance (variance-covariance matrix)
* Decompose into upper-triangular matrix


## Libraries
You need these.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(sp)
library(gstat)
library(geoR)
library(RColorBrewer)
library(classInt)
```


## Plot Data Locations

Step 1: Plot your geospatial data by assigning data appropriately to coordinates for spatial reference.

We're going to use the Meuse data, where four heavy metals (cadmium, copper, lead, zinc) were measured in the top soil in a flood plain along the river Meuse, along with a handful of covariates. The process governing heavy metal distribution seems that polluted sediment is carried by the river, and mostly deposited close to the river bank, and areas with low elevation. 

Here, we're going to look at zinc concentrations.

```{r meuse}
data(meuse)
#
coordinates(meuse) = c("x", "y")
#head(data)
#dim(data)

pal = brewer.pal(5,"Greens")
q5 = classIntervals(meuse$zinc, n=5, style="quantile")
q5Colors = findColours(q5,pal)
plot(meuse,col=q5Colors,pch=19,axes=T)
legend("topleft",fill=attr(q5Colors,"palette"),
	legend = names(attr(q5Colors,"table")),bty="n")
```

Let's take a look at the zinc concentrations more closely. 

```{r histzinc}
hist(meuse$zinc)
```

May not be normal (but lack of independance can cause things that are normal to look non-normal. Regardless, let's try a transformation.

```{r loghistzinc}
meuse$logzinc = log(meuse$zinc)
hist(meuse$logzinc)
```

Now look at the logzinc across locations.

```{r logzinc}
pal = brewer.pal(5,"Greens")
q5 = classIntervals(meuse$logzinc, n=5, style="quantile")
q5Colors = findColours(q5,pal)
plot(meuse,col=q5Colors,pch=19,axes=T)
legend("topleft",fill=attr(q5Colors,"palette"),
	legend = names(attr(q5Colors,"table")),bty="n")
```


```{r hdistance}
# Look at the correlation between points a distance h apart
hscat(logzinc~1,data=meuse, breaks=c(0:9)*100)

# How many points are there?
length(meuse$logzinc)

# How many pairs of points?
choose(155,2)
#
# Scatter plot of squared differences
plot(variogram(logzinc~1,meuse,cloud=T))
```

## Empirical Variogram


``` {r variogram}
# Calculate the empirical variogram
logzinc.vario = variogram(logzinc~1,meuse)
#
# Take a look at the contents of logzinc.vario
logzinc.vario
```




# Fit the model first by eye


## Fit the Actual Model
Now use these eye parameters to start the fit of model.



Plot the data, model and parameter estimates



## Predict
Let's make some predictions about the data, given the fitted model.



